---
title: "Explore algorithm divergence"
author: "Andrew bray"
date: "May 18, 2015"
output: html_document
---

```{r globalops, message = FALSE, error = FALSE}
library(knitr)
opts_chunk$set(message = FALSE, error = FALSE, comment=NA, fig.align="center", 
               echo = FALSE, fig.width=6, fig.height=6)
```

## Verify lines equivalence
```{r lines, message = FALSE, error = FALSE, fig.align="center", fig.width=6, fig.height=6}
library(dplyr)
library(devtools)
devtools::install_github("andrewpbray/lmmoptim", ref = "v0.0.1")
library("lmmoptim")

data(hmo)
data(hmo_states)
hmobig <- left_join(hmo, hmo_states, by = "state")

y <- hmobig$indPrem
n <- length(y)
mod <- lm(indPrem ~ expPerAdm + (region == "NE") + state, data = hmobig, x = TRUE)
x <- mod$x[, 1:3]
z <- mod$x[, -(1:3), drop = FALSE]
SigE <- diag(n)
SigS <- diag(44)

linesA <- findlines(x, z, y, SigE, SigS)

devtools::install_github("andrewpbray/lmmoptim", ref = "v0.0.0.9000")
library("lmmoptim")

linesB <- findlines(x, z, y, SigE, SigS)
```

The only difference is that the lines dataframe in the new code (linesA) has an 
extra column, "b", of all 1's, that the old code (linesB) does not have. They
are otherwise equivalent.


## Verify mlrebox equivalence
```{r mlrebox, message = FALSE, error = FALSE, fig.align="center", fig.width=6, fig.height=6}
mlreboxB <- with(linesB,
                makebox(lims.sigsqs = c(0, max(int.sigsqs[is.finite(int.sigsqs)])),
	                      lims.sigsqe = c(0, max(int.sigsqe[is.finite(int.sigsqe)])),
	                      status = rep("straddle", nrow(linesB)),
                        lines = linesB))

devtools::install_github("andrewpbray/lmmoptim", ref = "v0.0.1")
library("lmmoptim")

mlreboxA <- with(linesA,
                makebox(lims.sigsqs = c(0, max(int.sigsqs[is.finite(int.sigsqs)])),
	                      lims.sigsqe = c(0, max(int.sigsqe[is.finite(int.sigsqe)])),
	                      status = rep("straddle", nrow(linesA)),
                        lines = linesA))
```

The old code returns a box where the bounds are a dataframe with a row for every
term. For the new code, the terms have been summed over the column, leaving a
single lower and upper. This structure will persist into the algorithm as the box
list grows.


## Verify findf input equivalence
```{r input, message = FALSE, error = FALSE, eval = FALSE, fig.align="center", fig.width=6, fig.height=6}
lines <- linesA
startbox <- mlreboxA
eps <- 5
delE <- log(10)
delS <- log(10)
ratio <- TRUE
M <- 5
maxit <- 2

inactive <- list()
ninact <- 0
# check whether startbox is a box or a list of boxes
activeA <- ifelse(length(startbox) == 4 && identical(names(startbox),c("lims.sigsqe", 
                                                                       "lims.sigsqs", "status", "bounds")), list(startbox), startbox)
nact <- length(activeA)
lowbound <- -Inf
iter <- 0

killfunc <- function(box, lb, M, eps, delE, delS, ratio) {
  # lb is a lower bound on max f; it changes at each iteration. M, eps,
  # delE, delS stay constant throughout the iterations.
  cond.low <- box$bounds[2] < lb - M
  cond.eps <- diff(box$bounds) < eps
  cond.E <- ifelse(ratio, diff(log(box$lims.sigsqe)) < delE, diff(box$lims.sigsqe) <
                     delE)
  cond.S <- ifelse(ratio, diff(log(box$lims.sigsqs)) < delS, diff(box$lims.sigsqs) <
                     delS)
  return(cond.low || cond.eps || cond.E || cond.S)
}

low.act <- max(vapply(X = activeA, FUN = function(box) {
  box$bounds[1]
}, FUN.VALUE = 0.1))
lowbound <- max(lowbound, low.act)
kill <- vapply(X = activeA, FUN = killfunc, FUN.VALUE = TRUE, lb = lowbound,
               M = M, eps = eps, delE - delE, delS = delS, ratio = ratio)
ninact <- length(inactive)
        
lims.sigsqe = c(mean(activeA[[1]]$lims.sigsqe), activeA[[1]]$lims.sigsqe[2])
lims.sigsqs = c(mean(activeA[[1]]$lims.sigsqs), activeA[[1]]$lims.sigsqs[2])
statusA = activeA[[1]]$status

strad <- which(statusA == "straddle")
lines = lines[strad, ]

tmp1A <- lines$int.sigsqs + lines$slope * lims.sigsqe[1]
tmp1A[is.infinite(tmp1A)] <- NA

# value of the lines at the right side of the box
tmp2A <- lines$int.sigsqs + lines$slope * lims.sigsqe[2]
tmp2A[is.infinite(tmp2A)] <- NA

# where is the box relative to the lines?
above <- with ( lines, ifelse ( slope > -Inf, lims.sigsqs[1] > tmp1A, lims.sigsqe[1] > int.sigsqe ) )
below <- with ( lines, ifelse ( slope > -Inf, lims.sigsqs[2] < tmp2A, lims.sigsqe[2] < int.sigsqe ) )
stats <- rep("straddle", nrow(lines))
stats[above] <- "above"
stats[below] <- "below"
    
statusA[strad] <- stats

# we could get some of the bounds from the parent, but it's just as easy to
# recalculate them
boundsA <- getbounds(lims.sigsqe = lims.sigsqe, lims.sigsqs = lims.sigsqs,
                    status = statusA, lines = lines)


devtools::install_github("andrewpbray/lmmoptim", ref = "v0.0.0.9000")
library("lmmoptim")

lines <- linesB
startbox <- mlreboxB

inactive <- list()
ninact <- 0
# check whether startbox is a box or a list of boxes
ifelse ( length(startbox)==4 && 
           identical ( names(startbox),
                       c("lims.sigsqe", "lims.sigsqs", "status", "bounds")
           ),
         activeB <- list(startbox),
         activeB <- startbox
)
#active <- list()
#active[[1]] <- startbox
nact <- length(activeB)
lowbound <- -Inf
iter <- 0

# conditions under which a box becomes inactive
killfunc <- function ( box, lb, M, eps, delE, delS, ratio ) {
  # lb is a lower bound on max logRL; it changes at each iteration
  # M, eps, delE, delS stay constant throughout the iterations.
  cond.low <- sum ( box$bounds$upper < lb - M )
  cond.eps <- sum ( box$bounds$upper - box$bounds$lower ) < eps
  cond.E <- ifelse ( ratio,
                     diff ( log(box$lims.sigsqe) ) < delE,
                     diff ( box$lims.sigsqe ) < delE
  )
  cond.S <- ifelse ( ratio,
                     diff ( log(box$lims.sigsqs) ) < delS,
                     diff ( box$lims.sigsqs ) < delS
  )
  return ( cond.low || cond.eps || cond.E || cond.S )
}

low.act <- max ( vapply ( X = activeB,
                          FUN = function(box) { sum ( box$bounds$lower ) },
                          FUN.VALUE = 0.1
)
)
lowbound <- max ( lowbound, low.act )
kill <- vapply ( X = activeB,
                 FUN = killfunc,
                 FUN.VALUE = TRUE,
                 lb=lowbound, M=M,eps=eps, delE-delE, delS=delS, ratio=ratio
)
ninact <- length(inactive)

lims.sigsqe = c ( mean(activeB[[1]]$lims.sigsqe), activeB[[1]]$lims.sigsqe[2] )
lims.sigsqs = c ( mean(activeB[[1]]$lims.sigsqs), activeB[[1]]$lims.sigsqs[2] )
statusB = activeB[[1]]$status

strad <- which(statusB == "straddle")
lines = lines[strad,]
# value of the lines at the left side of the box
tmp1B <- with ( lines, ifelse ( is.finite(slope),
                               int.sigsqs + slope * lims.sigsqe[1],
                               NA
)
)
# value of the lines at the right side of the box
tmp2B <- with ( lines, ifelse ( is.finite(slope),
                               int.sigsqs + slope * lims.sigsqe[2],
                               NA
)
)
above <- with ( lines, ifelse ( slope > -Inf, lims.sigsqs[1] > tmp1B, lims.sigsqe[1] > int.sigsqe ) )
below <- with ( lines, ifelse ( slope > -Inf, lims.sigsqs[2] < tmp2B, lims.sigsqe[2] < int.sigsqe ) )
straddle <- !(above | below)

# sanity check
if ( any ( above & below ) ) print ( "a box can't be both above and below a line")

stats <- rep ( NA, nrow(lines) )
stats[above] <- "above"
stats[below] <- "below"
stats[straddle] <- "straddle"

statusB[strad] <- stats

boundsB <- getbounds ( lims.sigsqe = lims.sigsqe,
                      lims.sigsqs = lims.sigsqs,
                      status = statusB,
                      lines = lines
)
```

The problem appeared to be in the new version of getstatus(), which was not
treating the case of a slope of -Inf.


## Verify that bug is fixed

```{r checkbugfix, message = FALSE, error = FALSE, fig.align="center", fig.width=6, fig.height=6}
devtools::install_github("andrewpbray/lmmoptim")
library("lmmoptim")

system.time(
boxes.HH11A <- findf(lines = linesA, mlreboxA, eps = 5, delE = log(10), 
                     delS = log(10), ratio = TRUE, M = 5, maxit = 10)
)

p <- showboxes(boxes.HH11A)
p + scale_x_continuous(name = expression(sigma[e]^2)) +
    scale_y_continuous(name = expression(sigma[s]^2)) +
    theme_bw()

p <- showfunc(boxes.HH11A)
p + scale_x_log10(name = expression(sigma[e]^2),
                  breaks = c(100,300,1000,3000,10000)) +
    scale_y_log10(name = expression(sigma[s]^2),
                  breaks = c(30,100,300,1000,3000,10000)) +
    theme_bw()

devtools::install_github("andrewpbray/lmmoptim", ref = "v0.0.0.9000")
library("lmmoptim")

system.time(
boxes.HH11B <- fitlmm(lines = linesB, mlreboxB, eps = 5, delE = log(10), 
                     delS = log(10), ratio = TRUE, M = 5, maxit = 10)
)

p <- showboxes(boxes.HH11B)
p + scale_x_continuous(name = expression(sigma[e]^2)) +
    scale_y_continuous(name = expression(sigma[s]^2)) +
    theme_bw()

p <- showfunc(boxes.HH11B)
p + scale_x_log10(name = expression(sigma[e]^2),
                  breaks = c(100,300,1000,3000,10000)) +
    scale_y_log10(name = expression(sigma[s]^2),
                  breaks = c(30,100,300,1000,3000,10000)) +
    theme_bw()
```

Victory.

```{r checkbugfix, message = FALSE, error = FALSE, fig.align="center", fig.width=6, fig.height=6}
devtools::install_github("andrewpbray/lmmoptim")
library("lmmoptim")

boxes.HH11 <- findf(lines = linesA, mlreboxA, eps = 5, delE = log(10), 
                     delS = log(10), ratio = TRUE, M = 5, maxit = 10)

box2 <- with(linesA,
             makebox(lims.sigsqe = c(300, 1000),
                     lims.sigsqs = c(0, 1000),
	                   status = rep("straddle", nrow(linesA)),
                     lines = linesA))

boxes.HH11 <- findf(lines = linesA, box2, eps = 1, M = 10, maxit = 15)

linesA <- addprior(linesA, alpha_e = 1, beta_e = 0, alpha_s = 1.1, beta_s = .1)
box3 <- with(linesA,
             makebox(lims.sigsqe = c(300, 1000),
                     lims.sigsqs = c(0, 1000),
                     status = rep("straddle", nrow(linesA)),
                     lines = linesA))

boxes.HH11Bayes <- findf(lines = linesA, box3, eps = 1, M = 10, maxit = 20)


```


